---
title: "Class Feedback"
author: "Amy Hessl"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I requested feedback from the class on:

* Difficulty of most recent homework  
* What aspects of the class are going well
* What aspects of the class could be improved to help students learn more effectively  

#### Homework Difficulty

```{r diff, echo=-1}
HwDiff <- c(6.5, 7, 6, 7, 8, 7.5, 6, 5, 10, 7)
paste("Mean =", mean(HwDiff), "St Dev =", round(sd(HwDiff), digits=2))
```
Some students find the homwework relatively easy, others find it quite challenging, but the mean is spot on '7', which is appropriate for class.

#### What is going well...
* Lectures, open lab xxx  
* Bash script writing xx  
* Workload is manageable, expectations reasonable, pace is good xxx  
* Learning by doing, hands on  
* Code reading comprehension improving  

#### It would be great if you....
* More info on homework expectations/deliverables  
* Better notes on symbols  
* More detail on project expectations xx 
* Spend more time on each assignment xx
* More links on outside resources  
* Git is scary  
* More on collaborating with other scientists  
* Upload class content earlier  




