---
title: "Intro to R"
author: "Amy Hessl"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Introduction to R and RStudio
#### Questions  
*	How do I find my way around RStudio?  
*	How do I interact with R?  
*	How can I manage my environment?  
*	How do I install packages?  

#### Objectives
*	To gain familiarity with the various panes in the RStudio IDE (Integrated Development Environment)  
*	To gain familiarity with the buttons, short cuts and options in the RStudio IDE  
*	To understand variables and how to assign to them  
*	To be able to manage your workspace in an interactive R session  
*	To be able to use mathematical and comparison operations  
*	To be able to call __functions__  
*	Introduction to __package__ management  

#### Terms/Commands  
*	`ls()` to list the variables in your environment.  
*	`rm()` to delete objects in your environment.  
* `install.packages()` to install packages (libraries).  
*	`library(package_name)` to open a package in an R session.  
*	__Vector__ - an object in R where all the elements are of the same type (integer, logical, character, etc)  

### Motivation
Science is a multi-step process: once you’ve designed an experiment and collected data, the real fun begins! This lesson will teach you how to start this process using R and RStudio. We will begin with raw data, perform exploratory analyses, and learn how to plot results graphically. This example starts with the _gapminder-FiveYearData.csv_ dataset containing population information for many countries through time. 

Can you read the data into R? Can you plot the population for Senegal? Can you calculate the average income for countries on the continent of Asia? By the end of these lessons you will be able to do things like plot the populations for all of these countries in under a minute!

### Introduction to RStudio
RStudio is a free, open source R integrated development environment (IDE). It provides a built-in editor, works on all platforms (including on servers) and provides many advantages such as integration with version control and project management.

#### Basic layout
When you first open RStudio, you will be greeted by three panels:

1. The interactive R console (entire left)   
2. Environment/History (tabbed in upper right)  
3. Files/Plots/Packages/Help/Viewer (tabbed in lower right)    
Once you open files, such as R scripts, an editor panel will also open in the top left.  

#### Work flow within RStudio
There are three main ways one can work within RStudio.  

1.	Test and play within the interactive R console then copy code into a _.R_ file to run later.   
    +  This works well when doing small tests and initially starting off.  
    +  It quickly becomes laborious.  
2.	Start writing in a _.R_ file and use RStudio’s command `/` short cut to push current line, selected lines or modified lines to the interactive R console.  
    +  This is a great way to start; all your code is saved for later  
    +  You will be able to run the file you create from within RStudio or using R’s source() function.  
3.  Write your code in a _.Rmd_ rather than in _.R_.  This allows you to keep your notes, code, figures and tables all in a single document.  It's akin to a scientific notebook and can even be shared with a non-R user.  
    +   Takes a little longer  
    +   Helps you keep track of everything in one place  
    
#### Running Segments of Code 
RStudio offers you great flexibility in running code from within the editor window. There are buttons, menu choices, and keyboard shortcuts. To run the current line, you can: 

1.	Click on the "Run" button above the editor panel  
2.	Select “Run Lines” from the “Code” menu  
3.	Hit Ctrl-Enter in Windows or Linux or Command-Enter on OS X. (This shortcut can also be seen by selecting the run pull down.   
4.	To run a block of code, select it and then hit Run.  
5.	If you have modified a line of code within a block of code you have just run, there is no need to reselect the section and Run, you can use the next button along, Re-run the previous region. This will run the previous code block including the modifications you have made.  

### Introduction to R
Much of your time in R will be spent in the R interactive console. This is where you will run all of your code, and can be a useful environment to try out ideas before adding them to an _.R_ file or as a code block to _.Rmd_. This console in RStudio is the same as the one you would get if you typed in R in your command-line environment.

The first thing you will see in the R interactive session is a bunch of information, followed by a `>` and a blinking cursor. In many ways this is similar to the shell environment you learned about during the shell lessons: it operates on the same idea of a “Read, evaluate, print loop”: you type in commands, R tries to execute them, and then returns a result.

#### Using R as a calculator
The simplest thing you could do with R is do arithmetic:

```{r calc}
1 + 100
```

And R will print out the answer, with a preceding “[1]”. Don’t worry about this for now, we'll explain that later. For now think of it as indicating output.
Like bash, if you type in an incomplete command, R will wait for you to complete it:

```
1 +
+
```

Any time you hit return and the R session shows a `+` instead of a `>`, it means it’s waiting for you to complete the command. If you want to cancel a command you can simply hit `Esc` and RStudio will give you back the `>` prompt.

>Cancelling Commands
If you’re using R from the commandline instead of from within RStudio, you need to use Ctrl+C instead of Esc to cancel the command. This applies to Mac users as well!

Cancelling a command isn’t only useful for killing incomplete commands: you can also use it to tell R to stop running code (for example if its taking much longer than you expect), or to get rid of the code you’re currently writing.

When using R as a calculator, the order of operations is the same as you would have learned back in school.
From highest to lowest precedence:  

*	Parentheses: ( )  
*	Exponents: ^ or **  
*	Divide: /  
*	Multiply: *  
*	Add: +  
*	Subtract: -  
```{r}
3 + 5 * 2
```

Use parentheses to group operations in order to force the order of evaluation if it differs from the default, or to make clear what you intend.
```{r}
(3 + 5) * 2
```

This can get unwieldy when not needed, but clarifies your intentions. Remember that others may later read your code.
```
(3 + (5 * (2 ^ 2))) # hard to read
3 + 5 * 2 ^ 2       # clear, if you remember the rules
3 + 5 * (2 ^ 2)     # if you forget some rules, this might help
```

The text after each line of code is called a “comment”. Anything that follows after the hash symbol `#` is ignored by R when it executes code.
Really small or large numbers get a scientific notation:
```{r}
2/10000
```

Which is shorthand for “multiplied by 10^XX”. So `2e-4` is shorthand for `2 * 10^(-4)`.

You can write numbers in scientific notation too:
```{r}
5e3  # Note the lack of minus here
```

#### Mathematical functions
R has many built in mathematical functions. To call a function, we simply type its name, followed by open and closing parentheses. Anything we type inside the parentheses is called the function’s arguments:

```{r}
sin(1)  # trigonometry functions
```

```{r}
log(1)  # natural logarithm
```

```{r}
log10(10) # base-10 logarithm
```


```{r}
exp(0.5) # e^(1/2)
```

Don’t worry about trying to remember every function in R. You can simply look them up on Google, or if you can remember the start of the function’s name, use the tab completion in RStudio.
This is one advantage that RStudio has over R on its own, it has auto-completion abilities that allow you to more easily look up functions, their arguments, and the values that they take.

Typing a `?` before the name of a command will open the help page for that command. As well as providing a detailed description of the command and how it works, scrolling to the bottom of the help page will usually show a collection of code examples which illustrate command usage. We’ll go through an example later.

#### Comparing Things
We can also do comparison in R:

```{r}
1 == 1  # equality (note two equals signs, read as "is equal to")
```

```{r}
1 != 2  # inequality (read as "is not equal to")
```

```{r}
1 <  2  # less than
```

```{r}
1 <= 1  # less than or equal to
```

```{r}
1 > 0  # greater than
```

```{r}
1 >= -9 # greater than or equal to
```

### Variables and Assignment
We can store values in variables using the assignment operator `<-`, like this:

```{r}
x <- 1/40
```

Notice that assignment does not print a value. Instead, we stored it for later in something called a variable. x now contains the value 0.025:

```{r}
x
```

Look for the Environment tab in one of the panes of RStudio, and you will see that `x` and its value have appeared. Our variable `x` can be used in place of a number in any calculation that expects a number:

```{r}
log(x)
```

Notice also that variables can be reassigned:
```{r}
x <- 100
```

`x` used to contain the value 0.025 and now it has the value 100.
Assignment values can contain the variable being assigned to:
```{r}
x <- x + 1 #notice how RStudio updates its description of x on the top right tab
```

The right-hand side of the assignment can be any valid R expression. The right-hand side is fully evaluated before the assignment occurs.
Variable names can contain letters, numbers, underscores and periods. They __cannot__ start with a number nor contain spaces at all.

Different people use different conventions for long variable names, these include:  

*	periods.between.words  
*	underscores_between_words  
*	camelCaseToSeparateWords  

What you use is up to you but be consistent.
It is also possible to use the `=` operator for assignment:
```{r}
x = 1/40
```

But this is much less common among R users, because `=` is often used to designate values in an argument (like flags). So the recommendation is to use `<-`.

>__Challenge. Which of the following are valid R variable names? Try assigning something to them to see.__  

*	min_height  
*	max.height  
*	_age  
*	.mass  
*	MaxLength  
*	min-length  
*	2widths  
*	celsius2kelvin  

>__Challenge 2. What will be the value of each variable after each statement in the following program?__  

mass <- 47.5
age <- 122
mass <- mass * 2.3
age <- age - 20

>__Challenge. Run the code from the previous challenge and write a command to compare mass to age. Is mass larger than age?__   

#### Vectorization
One final thing to be aware of is that R is vectorized, meaning that variables and functions can have vectors as values. For example:  
```{r}
1:5
```


```{r}
2^(1:5)
```

```{r}
x <- 1:5
2^x
```
 
One can also use `rep` to repeat sequences:
 
```{r}
rep(1:5, 5)
```

This is incredibly powerful; we will discuss this further in an upcoming lesson.

>__Challenge. Make a variable called 'minutes' and repeat 1 thru 60 24 times.


### Managing your environment
There are a few useful commands you can use to interact with the R session.
`ls()` will list all of the variables and functions stored in the global environment (your working R session):

```{r}
ls()
```
You can also see what you have in your environment by looking at the Environment tab!

#### Hidden objects
Like in the shell, `ls` will hide any variables or functions starting with a “.” by default. To list all objects, type: 

```{r}
ls(all.names=TRUE)
```

Note here that we didn’t given any arguments to `ls`, but we still needed to give the parentheses to tell R to call the function. Also note the `=` here.

What is a .Random.seed?  Let's find out:
```{r, eval=F}
head(.Random.seed)
```

> __Challenge: What is a random seed?__  

Try typing `set.seed` into the help search window.

If we type `ls` by itself, R will print out the source code for that function!
```{r}
ls
```

You can use `rm` to delete objects you no longer need:

```{r}
rm(x)
```

If you have lots of things in your environment and want to delete all of them, you can pass the results of ls to the `rm` function:
```{r}
rm(list = ls())
```
In this case we’ve combined the two. Like the order of operations, anything inside the innermost parentheses is evaluated first, and so on.
In this case we’ve specified that the results of `ls` should be used for the list argument in `rm.` When assigning values to arguments by name, you must use the `=` operator!!
If instead we use `<-`, there will be unintended side effects, or you may get an error message:
```{r, eval=F}
rm(list <- ls())
```

#### Tip: Warnings vs. Errors
Pay attention when R does something unexpected! Errors, like above, are thrown when R cannot proceed with a calculation. Warnings on the other hand usually mean that the function has run, but it probably hasn’t worked as expected.
In both cases, the message that R prints out usually give you clues how to fix a problem. If you can't sort it out - try Google. More on getting help later.

>__Challenge. Clean up your working environment by deleting the mass and age variables from previous challenge.__ 

### R Packages
It is possible to add functions to R by writing a package, or by obtaining a package written by someone else. As of this writing, there are over 10,000 packages available on CRAN (the comprehensive R archive network). R and RStudio have functionality for managing packages: 

*	You can see what packages are installed by typing `installed.packages()`    
*	You can install packages by typing `install.packages("packagename")`, where packagename is the package name, in quotes.  
*	You can update installed packages by typing `update.packages()`  
*	You can remove a package with `remove.packages("packagename")`  
*	You can make a package available for use with `library(packagename)`  

R packages can also easily be installed from R Studio _Files/Plots/Packages_ etc frame.  

1.	Click on 'Packages'.      
2.	Click 'Install'.  
3.	Search for your package and install.  
4.	You will need to click in the library box to use the package in that session.  

What happens in the console when you install and run a package?


>__Challenge 5. Install the following packages: ggplot2, plyr, gapminder__ 

#### Key Points  
*	Use RStudio to write and run R programs.  
*	R has the usual arithmetic operators and mathematical functions.  
*	Use `<-` to assign values to variables.  
*	Use `ls()` to list the variables in a program.  
*	Use `rm()` to delete objects in a program.  
*	Use `install.packages()` to install packages (libraries).  
*	Use `library(package_name)` to open a package in each session.  

