---
title: "Base R Plotting"
subtitle: "GEOG 693g"
author: "Amy Hessl"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Motivation
Finally Sandra has her data all munged and ready to explore.  How can she get started making some figures that can be recreated later, just in case she needs to update her data?  How can she remake these figures for publication at the right resolution and size required by her journal?

### Questions
*How can I create simple X-Y plots in base R?  
*How can I create my own plots in base R with fine control?  
*How can I plot multiple graphs on a page?  
*How can I add marginal plots?  
*How can I plot graphs on top of each other (2 y-axes)?  

### Objectives
Develop some basic skills in X-Y plotting typical of exploratory data analysis.

### R Functions
* summary() - provides basic summary stats for every variable (column) in data frame
* attach() - database or object attached to the R search path which allows you to call variables without indicating the data frame first.  
* plot() - baseR plotting X-Y plots (scatter, line, stair-step etc)  
* color palette - a set of pre-selected colors.  Great for folks with limited artistic abilities.

### Base R Plotting is Fun
Using the original gapminder dataset, let's make a bubble plot of life expectancy, income and population.

```{r gapminder}
gap <- read.csv("../data/gapminder.csv")
summary(gap)
```

Let's extract the most recent year of data:
```{r gap07}
gap_07 <- gap[gap$year==2007,]
```

Let's make a simple scatter plot of lifeExp vs gdpPercap

```{r gapplot_simple, echo=TRUE}
plot(gap_07$lifeExp, gap_07$gdpPercap, type="p")
```

We need to transform these data!
```{r gapplot_lims, echo=TRUE}
summary(gap_07)
jXlim <- c(250, 50000)
jYlim <- c(39, 83)
attach(gap_07)
plot(lifeExp ~ gdpPercap, log = 'x', xlim = jXlim, ylim = jYlim)
```

Let's change those points to solid and scale them to population:
```{r gapplot_pop, echo=TRUE}
plot(lifeExp ~ gdpPercap, log = 'x', xlim = jXlim, ylim = jYlim, cex=pop/1e+08)
```

Better, how about we color the symbols based on continent? There are lots of color palletes created in R already - I stick with those! Here we display what a pallete of 5 colors would look like, then save the colors as a dataframe for future use. () allows us to display the dataframe as well as save it

```{r colorBrewer, echo=TRUE}
library(RColorBrewer)
display.brewer.pal(5, "Dark2")

(contCB <-
   with(gap_07,
        data.frame(continent = levels(continent),
                   color = I(brewer.pal(nlevels(continent), name = 'Dark2')))))   
```
With intention, this data.frame has a factor continent with the same name as the continent factor in jDat and with the same levels, in the same order. I am also using inhibit, I(), to protect the color-specifying hex strings from being converted to factor.


Now we can see how we might match the data in the color dataframe to the gapminder 'continent' variable.
```{r match, echo=TRUE}
head(data.frame(subset(gap_07, select = c(country, continent)),
           matchRetVal = match(gap_07$continent, contCB$continent)))
```
Now I can use the match() results to index onto the color variable of my color scheme. That creates the vector of colors I need for the col = argument of plot().

```{r gapplot_color, echo=TRUE}
plot(lifeExp ~ gdpPercap, log = 'x', xlim = jXlim, ylim = jYlim, 
     cex=pop/1e+08, pch=20,
     col = contCB$color[match(gap_07$continent, contCB$continent)],
     main = 'Life Expectancy vs GDP in 2007')
legend(x = 'bottomright', 
       legend = as.character(contCB$continent),
       col = contCB$color, pch = 20, bty = 'n', xjust = 1)
```
Nearly there - I don't like the way we scaled the population variable. Let's create a function that calculates the percentiles for each country and use that to scale.

```{r gaplot_pop, echo=TRUE}
perc.rank <- function(x) trunc(rank(x))/length(x)
gap_07$popPerc <- perc.rank(pop)
attach(gap_07)

plot(lifeExp ~ gdpPercap, log = 'x', xlim = jXlim, ylim = jYlim, 
     cex=(popPerc^2)*2, pch=20,
     col = contCB$color[match(gap_07$continent, contCB$continent)],
     main = 'Life Expectancy vs GDP in 2007',
     ylab='Life Expectancy', xlab='Log GDP/cap')
legend(x = 'bottomright', 
       legend = as.character(contCB$continent),
       col = contCB$color, pch = 20, bty = 'n', xjust = 1, y.intersp = 1)

```
Oh I don't know!  Come up with your own way to represent the values!

Finally, let's select some of the symbols and display a label for it.

```{r select, echo=TRUE}
plot(lifeExp ~ gdpPercap, log = 'x', xlim = jXlim, ylim = jYlim, 
     cex=(popPerc^2)*2, pch=20,
     col = contCB$color[match(gap_07$continent, contCB$continent)],
     main = 'Life Expectancy vs GDP in 2007',
     ylab='Life Expectancy', xlab='Log GDP/cap')
legend(x = 'bottomright', 
       legend = as.character(contCB$continent),
       col = contCB$color, pch = 20, bty = 'n', xjust = 1, y.intersp = 1.0)
identify(lifeExp ~ gdpPercap, labels=country)
```


### Add a Regression Line and Expression to Plot

First, create a regression model to add:
```{r linear model, message=FALSE, warning=FALSE}
splm <- lm(mpg~wt, data=mtcars)
summary.lm(splm)
adj.r <- round(summary.lm(splm)$adj.r.square, 4)
```

Next, use that formula to plot the line:
```{r abline with lm}
plot(mtcars$wt,mtcars$mpg, main="Scatterplot of wt vs. mpg")
abline(reg = splm) #abline(coef=coef(splm))
legend("topright", paste0("Adj.R2 = ", adj.r), bty='n')
```

### Combining Plots
R makes it easy to combine multiple plots into one overall graph, using either the par( ) or layout( ) function.

With the par( ) function, you can include the option mfrow=c(nrows, ncols) to create a matrix of nrows x ncols plots that are filled in by row. mfcol=c(nrows, ncols) fills in the matrix by columns.

##### _Before altering par, always save the defaults_
```{r default par, message=FALSE, warning=FALSE}
head(par())
def.par <- par()
```


#### Arrange Four Figures in One Plot 
```{r mtcars 2x2, message=FALSE, warning=FALSE}
par(mfrow=c(2,2))
plot(mtcars$wt, mtcars$mpg, main="Scatterplot of wt vs. mpg")
plot(mtcars$wt, mtcars$disp, main="Scatterplot of wt vs disp")
hist(mtcars$wt, main="Histogram of wt")
boxplot(mtcars$wt, main="Boxplot of wt")
```


#### Stack Three Figures as 3 rows and 1 column
```{r mtcars 3x1, message=FALSE, warning=FALSE}
par(mfrow=c(3,1), mar=c(2, 4, 1, 3)) 
hist(mtcars$wt)
hist(mtcars$mpg)
hist(mtcars$disp)
par <- par(def.par)
```


The layout( ) function has the form	layout(mat) where 
mat is a matrix object specifying the location of the N figures to plot.

#### Asymmetrical: One figure in row 1 and two figures in row 2
```{r mtcars 1x2, message=FALSE, warning=FALSE}
layout(matrix(c(1,1,2,3), 2, 2, byrow = TRUE))
hist(mtcars$wt)
hist(mtcars$mpg)
hist(mtcars$disp)
```

Optionally, you can include widths= and heights= options in the layout( ) function to control the size of each figure more precisely. These options have the form 
widths= a vector of values for the widths of columns
heights= a vector of values for the heights of rows.

Relative widths are specified with numeric values. Absolute widths (in centimetres) are specified with the lcm() function.

#### One figure in row 1 and two figures in row 2
row 1 is 1/3 the height of row 2
column 2 is 1/4 the width of the column 1 

```{r complex plot, message=FALSE, warning=FALSE}
layout(matrix(c(1,1,2,3), 2, 2, byrow = TRUE), 
  	widths=c(3,1), heights=c(1,2))
hist(mtcars$wt)
hist(mtcars$mpg)
hist(mtcars$disp)
```

See help(layout) for more details.

#### Creating a figure arrangement with fine control
In the following example, two box plots are added to scatterplot to create an enhanced graph by
setting the par(fig=c(x1, x2, y1, y2)) coordinates in the figure region.  Must use par(new=TRUE) in order to draw over same plot region each time.

Add boxplots to a scatterplot
```{r boxplots to a scatterplot}
par(fig=c(0,0.8,0,0.8), new=TRUE)
plot(mtcars$wt, mtcars$mpg, xlab="Car Weight",
  ylab="Miles Per Gallon")
par(fig=c(0,0.8,0.55,1), new=TRUE)
boxplot(mtcars$wt, horizontal=TRUE, axes=FALSE)
par(fig=c(0.65,1,0,0.8),new=TRUE)
boxplot(mtcars$mpg, axes=FALSE)
mtext("Enhanced Scatterplot", side=3, outer=TRUE, line=-3)
```

To understand this graph, think of the full graph area as going from (0,0) in the lower left corner to (1,1) in the upper right corner. The format of the fig= parameter is a numerical vector of the form c(x1, x2, y1, y2). The first fig= sets up the scatterplot going from 0 to 0.8 on the x axis and 0 to 0.8 on the y axis. The top boxplot goes from 0 to 0.8 on the x axis and 0.55 to 1 on the y axis. I chose 0.55 rather than 0.8 so that the top figure will be pulled closer to the scatter plot. The right hand boxplot goes from 0.65 to 1 on the x axis and 0 to 0.8 on the y axis. Again, I chose a value to pull the right hand boxplot closer to the scatterplot. You have to experiment to get it just right.

fig= starts a new plot, so to add to an existing plot use new=TRUE.

#### Plotting Multiple Time Series

```{r NASAGISS temp}
temps <- read.csv("https://data.giss.nasa.gov/gistemp/graphs/graph_data/Hemispheric_Temperature_Change/graph.csv")

head(temps[,1:4])
summary(temps[,1:4])

 par(oma=c(0,1,1,3), mar=c(5,4,5,3))
    plot(temps$Year, temps$NHem, type='l', axes=F, xlab='', ylab='', col="cyan")
    axis(1, seq(1880, 2020, by=20))
    mtext(side=1, "Year", line=2)
    axis(2, seq(-0.5, 1.3, by=0.5))
    mtext(side=2, "Temp Anom (C)", line=2)

    lines(temps$Year, temps$SHem, col="magenta")
```

```{r add CO2}
  
mlo <- read.table("ftp://aftp.cmdl.noaa.gov/products/trends/co2/co2_annmean_mlo.txt", skip=57)
colnames(mlo) <- c("Year", "CO2","Unc")

par(oma=c(0,1,1,3), mar=c(5,4,5,3))
    plot(temps$Year, temps$NHem, type='l', axes=F, xlab='', ylab='', col="cyan",
         xlim=c(1880, 2017))
    axis(1, seq(1880, 2020, by=20))
    mtext(side=1, "Year", line=2)
    axis(2, seq(-0.5, 1.3, by=0.5))
    mtext(side=2, "Temp Anom (C)", line=2)

    lines(temps$Year, temps$SHem, col="magenta")
    
par(new=TRUE)
    plot(mlo$Year, mlo$CO2, lty=3, type='l', col="grey10", axes=F, xlab='', 
         ylab='', xlim=c(1880, 2017))
    axis(4, seq(325, 400, by=25))
    mtext(side=4, "CO2 (ppm)", line=2)
    
    legend(x="bottomright", legend=c("NHem", "SHem", "CO2"),  
           col= c("cyan", "magenta", "grey10"), lty=c(1,1,3), bty="n", xjust=-2)
```

#### Publication Quality Files
Should be overwritten each time to prevent too many versions. Settings can easily be changed to write to pdf() or other format depending on requirements for publication.


```{r write to a plot file, eval=FALSE}

ppi=300
png("/products/NASAGISS.png", width=8*ppi, height=6*ppi, res=ppi)

par(oma=c(0,1,1,3), mar=c(5,4,5,3))
    plot(temps$Year, temps$NHem, type='l', axes=F, xlab='', ylab='', col="cyan",
         xlim=c(1880, 2017))
    axis(1, seq(1880, 2020, by=20))
    mtext(side=1, "Year", line=2)
    axis(2, seq(-0.5, 1.3, by=0.5))
    mtext(side=2, "Temp Anom (C)", line=2)

    lines(temps$Year, temps$SHem, col="magenta")
    
par(new=TRUE)
    plot(mlo$Year, mlo$CO2, lty=3, type='l', col="grey10", axes=F, xlab='', 
         ylab='', xlim=c(1880, 2017))
    axis(4, seq(325, 400, by=25))
    mtext(side=4, "CO2 (ppm)", line=2)
    
    legend(x="bottomright", legend=c("NHem", "SHem", "CO2"),  
           col= c("cyan", "magenta", "grey10"), lty=c(1,1,3), bty="n", xjust=-2)
dev.off()
```

Always reset the graphical parameters to the default (or use the paintbrush button in Rstudio)
```{r reset default par(), message=FALSE, warning=FALSE}
par <- par(def.par)
```

#### Key Points
*	Base R allows for very fine control of plots.  But can take some time to develop.

#### __Acknowledgments__
_These pages were derived in part from: https://www.statmethods.net/advgraphs/layout.html_