---
title: "Dataframes"
author: "Amy Hessl"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Motivation
The star of the R show is the data frame.  Most of our data are stored in flat files where every row is an observation and every column is a variable.  R was designed with  this data format in mind.  Being able to manipulate data in data frames is an esssential skill in any science.  While most of what we will do today is easy in excel, R allows you to:  
1) Track exactly how the raw data were altered.  
2) Reproduce your data on any number of similar data frames.  
3) Go back and edit your code to produce different outputs.  


#### Questions  
*	How can I manipulate a data frame?
* How can I access particular rows or columns?
* How can I remove particular rows or columns?
* How can I append two similar dataframes together?  

#### Objectives  
*	Add and remove rows or columns.  
*	Remove rows with *NA* values.  
*	Append (concatenate) two data frames.  
*	Understand how to use a factor.  
*	Convert a factor to a character vector and vice versa.  
*	Display basic properties of data frames including size and class of the columns, names, and first few rows.  

#### Terms/Commands 
*	`cbind()` to add a new column to a data frame.
*	`rbind()` to add a new row to a data frame.
*	Remove rows from a data frame using [-x,], where x is the row to remove.
*	`na.omit()` to remove rows from a data frame with NA values.
*	`length()` to get the number of rows in a data frame.

At this point, you’ve seen it all: in the last lesson, we toured all the basic data types and data structures in R. Everything you do i R will be a manipulation of those basic building blocks. But most of the time, the star of the show is the data frame. In this lesson, we’ll learn a few more things about working with data frames.

##### Data
Be sure you have *feline-data.csv* loaded.

```{bash, echo=FALSE}
echo "coat,weight,likes_string
calico,2.1,1
black,5.0,0
tabby,3.2,1" > ../data/feline-data.csv
```
```{r, echo=FALSE}
cats <- read.csv(file="../data/feline-data.csv")
```

#### Adding columns and rows in data frames 
We already learned that the columns of a data frame are vectors, so that our data are consistent in type throughout the columns. As such, if we want to add a new column, we can start by making a new vector:
```{r}
age <- c(2, 3, 5)
cats
```

        1
We can then add this as a column via:
```{r, echo=T}
cbind(cats, age)
```
    
Note that if we tried to add a vector of ages with a different number of entries than the number of rows in the dataframe, it would fail:
```{r, eval=F}
age <- c(2, 3, 5, 12)
cbind(cats, age)
```

Why didn’t this work? Of course, R wants to see one element in our new column for every row in the table:
```{r}
nrow(cats)
```
```{r}
length(age)
```

So for it to work we need to have `nrow(cats) = length(age)`. Let’s overwite the content of cats with our new data frame.
```{r}
age <- c(2, 3, 5)
cats <- cbind(cats, age)
```

Now how about adding rows? We already know that the rows of a data frame are lists, therefore adding a new row to a data frame requires that it be a list (it might have different types of data), but we have a problem!
```{r, eval=F}
newRow <- list("tortoiseshell", 3.3, TRUE, 9)
cats <- rbind(cats, newRow)
```

#### Factors
Here is another thing to look out for: in a factor, each different value represents what is called a level. In our case, the factor “coat” has 3 levels: “black”, “calico”, and “tabby”. R will only accept values that match one of the levels. If you add a new value, it will become NA.

The warning is telling us that we unsuccessfully added “tortoiseshell” to our coat factor, but 3.3 (a numeric), TRUE (a logical), and 9 (a numeric) were successfully added to weight, likes_string, and age, respectively, since those variables are not factors. To successfully add a cat with a “tortoiseshell” coat, we must add “tortoiseshell” as a level of the factor:
```{r}
levels(cats$coat)
```
```{r}
levels(cats$coat) <- c(levels(cats$coat), "tortoiseshell")
cats <- rbind(cats, list("tortoiseshell", 3.3, TRUE, 9))
```
Alternatively, we can change a factor into a character vector; we lose the handy categories of the factor, but we can subsequently add any word we want to the column without babysitting the factor levels:
```{r}
str(cats)
```
```{r}
cats$coat <- as.character(cats$coat)
str(cats)
```

>__Challenge:__ Let’s imagine that 1 human year is equivalent to 7 cat years. 

1.	Create a vector called human_age by multiplying cats$age by 7.  
2.	Convert human_age to a factor.  
3.	Convert human_age back to a numeric vector using the `as.numeric()` function. Now divide it by 7 to get the original ages back.
***

#### Removing Rows
We now know how to add rows and columns to our data frame in R—but in our first attempt to add a “tortoiseshell” cat to the data frame we have accidentally added a garbage row:
```{r}
cats
```
     
We can ask for a data frame minus this offending row:
```{r}
cats[-4, ]
```
           
Notice the comma with nothing after it to indicate that we want to drop the entire fourth row.

Note: we could also remove both new rows at once by putting the row numbers inside of a vector: 
```{r}
cats[c(-4,-5), ]
```

Alternatively, we can drop all rows with NA values:
```{r}
na.omit(cats)
```
           
Let’s reassign the output to cats, so that our changes will be permanent:

```{r}
cats <- na.omit(cats)
```

#### Removing Columns
We can also remove columns in our data frame. What if we want to remove the column “age”. We can remove it in two ways, by variable number or by index.
```{r}
cats[,-4]
```
 
Notice the comma with nothing before it, indicating we want to keep all of the rows.
Alternatively, we can drop the column by using an index based on the column names. 

```{r}
drop <- names(cats) %in% c("age")
(cats <- cats[,!drop])
```
>__Challenge:__ What kind of vector is `drop`? What does it represent?

What is the advantage of using the row name index as opposed to the position of 'age' `[,-4]`?
***

#### Appending to a Data Frame
The key to remember when adding data to a data frame is that columns are vectors and rows are lists. We can also glue two data frames together with `rbind`:
```{r}
cats <- rbind(cats, cats)
cats
```
            
But now the row names are unnecessarily complicated. We can remove the rownames, and R will automatically re-name them sequentially:
```{r}
rownames(cats) <- NULL
cats
```
          
>__Challenge:__You can create a new data frame right from within R with the following syntax:

```{r}
df <- data.frame(id = c("a", "b", "c"),
                 x = 1:3,
                 y = c(TRUE, TRUE, FALSE),
                 stringsAsFactors = FALSE)
```

Make a data frame that holds the following information for yourself:
*	first name  
*	last name  
*	lucky number  
Then use `rbind` to add an entry for the people sitting beside you. Finally, use `cbind` to add a column with each person’s answer to the question, “Is it time for donuts?”

***

#### A Realistic Example
So far, you have seen the basics of manipulating data frames with our cat data; now let’s use those skills to digest a more realistic dataset. Let’s read in the gapminder dataset.  If you downloaded it previously, you can read it in from your data directory in your current project:
```{r, eval=FALSE}
gapminder <- read.csv("../data/gapminder_data.csv")
```
Or you can get it from the gapminder package:
```{r}
gapminder <- gapminder::gapminder
```

#### Miscellaneous Tips
*	Another type of file you might encounter are tab-separated value files (.tsv). To specify a tab as a separator, use "\\t" or read.delim().  
*	Files can also be downloaded directly from the Internet into a local folder of your choice onto your computer using the download.file function. The read.csv function can then be executed to read the downloaded file from the download location, for example,
```{r}
download.file("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv", destfile = "../data/gapminder_data.csv")
```
Then read it in:
```{r}
gapminder <- read.csv("../data/gapminder_data.csv")
```
*	Alternatively, you can also read in files directly into R from the Internet by replacing the file paths with a web address in `read.csv`. One should note that in doing this no local copy of the csv file is first saved onto your computer. For example:
```{r}
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv")
```
*	You can read directly from excel spreadsheets without converting them to plain text first by using the `readxl` package.  

Let’s investigate gapminder a bit; the first thing we should always do is check out what the data looks like with `str`:
```{r}
str(gapminder)
```

We can also interrogate the data frame for information about its dimensions; remembering that `str(gapminder)` said there were 1704 observations of 6 variables in gapminder, what do you think the following will produce, and why?
```{r}
length(gapminder)
```

A fair guess would have been to say that the length of a data frame would be the number of rows it has (1704), but this is not the case; remember, a data frame is a list of vectors and factors:

```{r}
typeof(gapminder)
```

When length gave us 6, it’s because gapminder is built out of a list of 6 columns. To get the number of rows and columns in our dataset, try:
```{r}
nrow(gapminder)
```
```{r}
ncol(gapminder)
```
Or, both at once:
```{r}
dim(gapminder)
```

We’ll also likely want to know what the titles of all the columns are, so we can ask for them later:
```{r}
colnames(gapminder)
```

At this stage, it’s important to ask ourselves if the structure R is reporting matches our intuition or expectations; do the basic data types reported for each column make sense? If not, we need to sort any problems out now before they turn into bad surprises down the road, using what we’ve learned about how R interprets data, and the importance of strict consistency in how we record our data.
Once we’re happy that the data types and structures seem reasonable, it’s time to start digging into our data proper. Check out the first few lines:
```{r}
head(gapminder)
```

>__Challenge__ Check top bottom and middle.

It’s good practice to also check the last few lines of your data and some in the middle. How would you do this?
Searching for ones specifically in the middle isn’t too hard but we could simply ask for a few lines at random. How would you code this?
Tip: There are several ways to achieve this.  

***


>__Challenge__ Source a R Script

Go to file -> new file -> R script, and write an R script to load in the gapminder dataset. Put it in the scripts/ directory and add it to version control.
Run the script using the `source()` function, using the file path as its argument (or by pressing the “source” button in RStudio).


Read the output of `str(gapminder)` again; this time, use what you’ve learned about factors, lists and vectors, as well as the output of functions like colnames and dim to explain what everything that str prints out for gapminder means. If there are any parts you can’t interpret, discuss with your neighbors!

***

#### Key Points 

*	Use `cbind()` to add a new column to a data frame.
*	Use `rbind()` to add a new row to a data frame.
*	Remove rows from a data frame.
*	Use `na.omit()` to remove rows from a data frame with NA values.
*	Use `levels()` and `as.character()` to explore and manipulate factors.
*	Use `str()`, `nrow()`, `ncol()`, `dim()`, `colnames()`, `rownames()`, `head()`, and `typeof()` to understand the structure of a data frame.
*	Read in a *csv* file using `read.csv()`.
*	Understand what `length()` of a data frame represents.


#### __Acknowledgments__
_These pages were derived in part from:Benjamin Soltoff's MACS 30500 - [Computing for the Social Sciences](https://cfss.uchicago.edu/) at University, and from [Software Carpentry](software-carpentry.org), licensed under the CC BY-NC 3.0 Creative Commons_
