---
title: "Control Flow: Functions"
subtitle: "GEOG 693g"
author: "Amy Hessl"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Motivation
Jonathan has a complicated set of calculations he needs to make on every data sheet.  How can he create his own function to make those calculations, simply by calling the function datCalc()?

#### Questions  
*	How can I write a new function in R?  
*	Define a function that takes arguments.  
*	Return a value from a function.  
*	Check argument conditions with stopifnot() in functions.  
*	Test a function.  
*	Set default values for function arguments.  
*	Explain why we should divide programs into small, single-purpose functions.  

In this lesson, we’ll learn how to write a function so that we can repeat several operations with a single command.

### What is a function?
Functions gather a sequence of operations into a whole, preserving it for ongoing use. Functions provide:  
*	a name we can remember and invoke it by  
*	relief from the need to remember the individual operations  
*	a defined set of inputs and expected outputs  
*	connections to the larger programming environment for example by sharing your function on GitHub!  

As the basic building block of most programming languages, user-defined functions constitute “programming” as much as any single abstraction can. If you have written a function, you are a computer programmer.

#### Defining a function
Let’s open a new R script file in a new directory called _functions/_  and call it _functions-lesson.R_.  Type in the following:

```{r}
my_sum <- function(a, b) {
  the_sum <- a + b
  return(the_sum)
}
```
#### How can we use a function?  
We must source it first (read it into R).  You can Run it from RStudio, enter it at the command line or source it:
```{r}
source("functions/functions-lesson.R")
my_sum(2, 3)
```
Check out what happened in your Global Environment.  What new object type comes up?

Let’s define a new function fahr_to_kelvin() that converts temperatures from Fahrenheit to Kelvin:
```{r}
fahr_to_kelvin <- function(temp) {
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}
```
*	We define fahr_to_kelvin() by assigning it to the output of function.    
*	The list of argument names are contained within parentheses (). In this case (temp).    
*	Next, the body of the function–the statements that are executed when it runs–is contained within curly braces ({}).    
*	The statements in the body are indented by two spaces. This makes the code easier to read but does not affect how the code operates.    

It is useful to think of creating functions like writing a cookbook. First you define the “ingredients” that your function needs. In this case, we only need one ingredient to use our function: “temp”. After we list our ingredients, we then say what we will do with them, in this case, we are taking our ingredient and applying a set of mathematical operators to it.
When we call the function, the values we pass to it as arguments are assigned to those variables so that we can use them inside the function. Inside the function, we use a *return statement* to send a result back to whoever asked for it. 

#### Return Statement
One feature unique to R is that the return statement is not required. R automatically returns whichever variable is on the last line of the body of the function. But for clarity, we will explicitly define the return statement.
Let’s try running our function. Calling our own function is no different from calling any other function:
```{r}
# freezing point of water
fahr_to_kelvin(32)
```
# boiling point of water
```{r}
fahr_to_kelvin(212)
```

>__Challenge__ 

Write a function called kelvin_to_celsius() that takes a temperature in Kelvin and returns that temperature in Celsius.
Hint: To convert from Kelvin to Celsius you subtract 273.15

***

#### Combining functions
The real power of functions comes from mixing, matching and combining them into ever-larger chunks to get the effect we want.
Let’s define two functions that will convert temperature from Fahrenheit to Kelvin, and Kelvin to Celsius:

```{r}
fahr_to_kelvin <- function(temp) {
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}

kelvin_to_celsius <- function(temp) {
  celsius <- temp - 273.15
  return(celsius)
}
```

>__Challenge__

Define a function to convert directly from Fahrenheit to Celsius, by reusing the two functions above (or using your own functions if you prefer).
***

### Interlude: Defensive Programming
Now that we’ve begun to appreciate how writing functions provides an efficient way to make R code re-usable and modular, we should note that it is important to ensure that functions only work in their intended use-cases. Checking function parameters is related to the concept of defensive programming. Defensive programming encourages us to frequently check conditions and throw an error if something is wrong. These checks are referred to as *assertion statements* because we want to assert some condition is TRUE before proceeding. They make it easier to debug because they give us a better idea of where the errors originate.

#### Checking conditions with `stopifnot()`
Let’s start by re-examining fahr_to_kelvin(), our function for converting temperatures from Fahrenheit to Kelvin. It was defined like so:
```
fahr_to_kelvin <- function(temp) {
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}
```

For this function to work as intended, the argument `temp` must be a numeric value; otherwise, the mathematical procedure for converting between the two temperature scales will not work. To create an error, we can use the function `stop()`. For example, since the argument `temp` must be a numeric vector, we could check for this condition with an if statement and throw an error if the condition was violated. We could augment our function above like so:
```{r}
fahr_to_kelvin <- function(temp) {
  if (!is.numeric(temp)) {
    stop("temp must be a numeric vector.")
  }
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}
```
If we had multiple conditions or arguments to check, it would take many lines of code to check all of them. Luckily R provides the convenience function `stopifnot()`. We can list as many requirements that should evaluate to TRUE; `stopifnot()` throws an error if it finds one that is FALSE. Listing these conditions also serves a secondary purpose as extra documentation for the function. 
Let’s try out defensive programming with `stopifnot()` by adding assertions to check the input to our function `fahr_to_kelvin()`.
We want to assert the following: temp is a numeric vector. We may do that like so:

```{r}
fahr_to_kelvin <- function(temp) {
  stopifnot(is.numeric(temp))
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}
```
It still works when given proper input.
```{r}
# freezing point of water
fahr_to_kelvin(temp = 32)
```

But fails instantly if given improper input.
```{r, eval=FALSE}
# Metric is a factor instead of numeric
fahr_to_kelvin(temp = as.factor(32))
```
>__Challenge__

Use defensive programming to ensure that our `fahr_to_celsius()` function throws an error immediately if the argument `temp` is specified inappropriately.
***

####  More on combining functions
Now, we’re going to define a function that calculates the Gross Domestic Product of a nation from the data available in our gapminder dataset:

```{r}
# Takes a dataset and multiplies the population column
# with the GDP per capita column.
calcGDP <- function(dat) {
  gdp <- dat$pop * dat$gdpPercap
  return(gdp)
}
```
We define calcGDP() by assigning it to the output of function. The list of argument names are contained within parentheses. Next, the body of the function – the statements executed when you call the function – is contained within curly braces ({}).

```{r}
gapminder <- read.csv("../data/gapminder.csv")
calcGDP(head(gapminder))
```
That’s not very informative. Let’s add some more arguments so we can extract that per year and country.
```{r}
# Takes a dataset and multiplies the population column
# with the GDP per capita column.
calcGDP <- function(dat, year=NULL, country=NULL) {
  if(!is.null(year)) {
    dat <- dat[dat$year %in% year, ]
  }
  if (!is.null(country)) {
    dat <- dat[dat$country %in% country,]
  }
  gdp <- dat$pop * dat$gdpPercap

  new <- cbind(dat, gdp=gdp)
  return(new)
}
```
If you’ve been writing these functions down into a separate R script (a good idea!), you can load in the functions into our R session by using the source() function:
```{r}
source("functions/functions-lesson.R")
```

Ok, so there’s a lot going on in this function now. In plain English, the function now subsets the provided data by year if the year argument isn’t empty, then subsets the result by country if the country argument isn’t empty. 
Then it calculates the GDP for whatever subset emerges from the previous two steps. The function then adds the GDP as a new column to the subsetted data and returns this as the final result. You can see that the output is much more informative than a vector of numbers.
Let’s take a look at what happens when we specify the year:
```{r}
head(calcGDP(gapminder, year=2007))
```      

Or both:
```{r}
calcGDP(gapminder, year=2007, country="Australia")
```
   
Let’s walk through the body of the function:

`calcGDP <- function(dat, year=NULL, country=NULL) {`
Here we’ve added two arguments, year, and country. We’ve set default arguments for both as NULL using the = operator in the function definition. This means that those arguments will take on those values **unless** the user specifies otherwise.
```
  if(!is.null(year)) {
    dat <- dat[dat$year %in% year, ]
  }
  if (!is.null(country)) {
    dat <- dat[dat$country %in% country,]
  }
```

Here, we check whether each additional argument is set to null, and whenever they’re not null overwrite the dataset stored in _dat_ with a subset given by the non-null argument.
I did this so that our function is more flexible for later. We can ask it to calculate the GDP for:   

*	The whole dataset;  
*	A single year;  
*	A single country;  
*	A single combination of year and country.  
By using %in% instead, we can also give multiple years or countries to those arguments!

#### Pass by value
Functions in R almost always make copies of the data to operate on inside of a function body. When we modify dat inside the function we are modifying the copy of the gapminder dataset stored in dat, not the original variable we gave as the first argument.
This is called “pass-by-value” and it makes writing code much safer: you can always be sure that whatever changes you make within the body of the function, stay inside the body of the function.

#### Function scope
Another important concept is scoping: any variables (or functions!) you create or modify inside the body of a function only exist for the lifetime of the function’s execution. When we call calcGDP(), the variables dat, gdp and new only exist inside the body of the function. Even if we have variables of the same name in our interactive R session, they are not modified in any way when executing a function. After running the function, look at your Global Environment.  Are there any new objects?
```
  gdp <- dat$pop * dat$gdpPercap
  new <- cbind(dat, gdp=gdp)
  return(new)
}
```

Finally, we calculated the GDP on our new subset, and created a new data frame with that column added (but note this dataframe is not saved, only returned to console). This means when we call the function later we can see the context for the returned GDP values, which is much better than in our first attempt where we got a vector of numbers.

>__Challenge__

Test out your GDP function by calculating the GDP for New Zealand in 1987. How does this differ from New Zealand’s GDP in 1952?
***

>__Challenge__

The paste() function can be used to combine text together, e.g:
```{r}
best_practice <- c("Write", "programs", "for", "people", "not", "computers")
paste(best_practice, collapse=" ")
```

Write a function called fence() that takes two vectors as arguments, called text and wrapper, and prints out the text wrapped with the wrapper:
`fence(text=best_practice, wrapper="***")`

Note: the paste() function has an argument called sep, which specifies the separator between text. The default is a space: “ “. The default for paste0() is no space “”.
***

#### Testing and documenting
It’s important to both test functions and document them: documentation helps you, and others, understand what the purpose of your function is, and how to use it, and it's important to make sure that your function actually does what you think.
When you first start out, your workflow will probably look a lot like this:
1.	Write a function  
2.	Comment parts of the function to document its behavior  
3.	Load in the source file  
4.	Experiment with it in the console to make sure it behaves as you expect  
5.	Make any necessary bug fixes  
6.	Rinse and repeat.  
Formal documentation for functions, written in separate _.Rd_ files, gets turned into the documentation you see in help files. The roxygen2 package allows R coders to write documentation alongside the function code and then process it into the appropriate .Rd files. You will want to switch to this more formal method of writing documentation when you start writing more complicated R projects.
Formal automated tests can be written using the testthat package.

#### Key Points 
*	Use function to define a new function in R.  
*	Use parameters to pass values into functions.  
*	Use stopifnot() to flexibly check function arguments in R.  
*	Load functions into programs using source().  

####mFurther Reading:
You can learn more by reading the [R Language Definitions](https://cran.r-project.org/doc/manuals/r-release/R-lang.html) or this [chapter](https://adv-r.hadley.nz/environments.html) from [Advanced R Programming](https://adv-r.hadley.nz/) by Hadley Wickham.

#### __Acknowledgments__
_These pages were derived in part from:Benjamin Soltoff's MACS 30500 - [Computing for the Social Sciences](https://cfss.uchicago.edu/) at University, and from [Software Carpentry](software-carpentry.org), licensed under the CC BY-NC 3.0 Creative Commons_
