---
title: "Getting Help in R"
author: "Amy Hessl"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Questions
*	How can I get help in R?  

#### Objectives
*	Be able read R help files for functions and special operators.  
*	Be able to use CRAN task views to identify packages to solve a problem.  
*	To be able to seek help from your peers.  

#### Reading Help files
R, and every package, provide help files for functions. To search for help on a function from a specific function that is in a package loaded into your namespace (your interactive R session):
`?function_name`
`help(function_name)`

This will load up a help page in RStudio (or as plain text in R by itself). Note you can also search for help in the RStudio help page itself.

#### Unpacking a R Help Page
Using the help search bar, you can find the `cat` {base} documentation.  You can also find it on the official R-manual page:
[Concatenate and Print](ttps://stat.ethz.ch/R-manual/R-devel/library/base/html/cat.html)

Help pages are all formatted the same way and include crucial info:  

* Function {package}  
* Title  
* Description: An extended description of what the function does.  
* Usage: The arguments of the function and their default values.   
* Arguments: An explanation of the data each argument is expecting.   
* Details - greater detail than description, including the kinds of objects allowed (atomic vectors, names, characters, data frames, lists etc.)  
* Value - this is the output of the function    
* Authors   
* See Also: Any related functions you might find useful. 
* Examples - these are completely self-contained examples (data, function and arguments included).  It is strongly recommended to try them before using a new function.  

#### Help for Functions
One of the most daunting aspects of R is the large number of functions available. It would be prohibitive, if not impossible to remember the correct usage for every function you use. Luckily, the help files mean you don’t have to!It's easy to get help either from the Console or from the search bar under "Help":
```{r}
?log
```

#### Special Operators
To seek help on special operators, use quotes:
```{r}
?"+"
```

#### When you kind of remember the function
If you’re not sure what package a function is in, or how it’s specifically spelled you can do a fuzzy search in the search bar under "Help".  Try searching for "cat".

***  

> __Challenge `paste`__

Look at the help for the `paste` function. What is the difference between the `sep` and `collapse` arguments?

#### When you have no idea where to begin
If you don’t know what function or package you need to use CRAN Task Views is a specially maintained list of packages grouped into fields. This can be a very good starting point.  

#### Getting help on packages
Many packages come with “vignettes”: tutorials and extended example documentation. Without any arguments, `vignette()` will list all vignettes for all installed packages; `vignette(package="package-name")` will list all available vignettes for package-name, and `vignette("vignette-name")` will open the specified vignette.
If a package doesn’t have any vignettes, you can usually find help by typing `help("package-name")`.  

> __Challenge: Install and Get Help for Packages__

1. Try finding vignettes for `ProjectTemplate`.  
2. Try installing `dplR` and opening a vignette.  

#### When your code doesn’t work: seeking help from your peers
If you’re having trouble using a function, 9 times out of 10, the answers you are seeking have already been answered on [Stack Overflow](https://stackoverflow.com/). You can search using the [r] tag.
If you can’t find the answer, there are a few useful functions to help you ask a question from your peers:

```?dput```

Will dump the data you’re working with into a format so that it can be copy and pasted by anyone else into their R session.

```{r}
sessionInfo()
```

Will print out your current version of R, as well as any packages you have loaded. This can be useful for others to help reproduce and debug your issue.

> __Challenge__  

Use help to find a function (and its associated arguments) that you could use to read data into R from a csv file in which columns are delimited with "\\t" (tab) and the decimal point is a “.” (period).

This check for decimal separator is important, especially if you are working with international colleagues, because different countries have different conventions for the decimal point (i.e. comma vs period). hint: use `??csv` to lookup csv related functions.

#### Other ports of call
* [Quick R](https://www.statmethods.net/)  
* [RStudio cheat sheets](https://rstudio.com/resources/cheatsheets/)  
* [Cookbook for R](http://www.cookbook-r.com/)  
* [How to make reproducible example in R](https://stackoverflow.com/questions/5963269/how-to-make-a-great-r-reproducible-example)  

#### Key Points
* Use `help()` to get online help in R.  
* Use [Stack Overflow](https://stackoverflow.com/) for more challenging problems.  
* Use `?dput` and `sessionInfo` to help document your problem.  

