---
title: "Control Flow: If/Else"
subtitle: "GEOG 693g"
author: "Amy Hessl"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Motivation
Elizabeth has a large number of files from a Hobo weather station.  Some of those files are in Celsius and some are in Fahrenheit.  She needs to convert all of those file to Kelvin.  She thinks an if/else statement is the way to get started.

#### Questions
*	How can I make data-dependent choices in R?
*	How can I repeat operations in R?

#### Objectives
*	Write conditional statements with if() and else().

Often when we’re coding we want to control the flow of our actions. This can be done by setting actions to occur only if a condition or a set of conditions are met. Alternatively, we can also set an action to occur a particular number of times.
There are several ways you can control flow in R. For conditional statements, the most commonly used approaches are the constructs if() and if()...else():  
**if**
```
if (condition is true) {
  perform action
}
```
***if...else***
```
if (condition is true) {
  perform action
} else {  # that is, if the condition is false,
  perform alternative action
}
```

Say, for example, that we want R to print a message if a variable x has a particular value:
```{r}
x <- 8

if (x >= 10) {
  print("x is greater than or equal to 10")
}
```

We get no response in the console because x is not greater than 10! To print a different message for numbers less than 10, we can add an else statement.
```{r}
x <- 8

if (x >= 10) {
  print("x is greater than or equal to 10")
} else {
  print("x is less than 10")
}
```

You can also test multiple conditions by using `else if ()`:  
```{r}
x <- 8

if (x >= 10) {
  print("x is greater than or equal to 10")
} else if (x > 5) {
  print("x is greater than 5, but less than 10")
} else {
  print("x is less than 5")
}
```
Important: when R evaluates the condition inside if() statements, it is looking for a logical element, i.e., TRUE or FALSE. This can cause some headaches for beginners. For example:
```{r}
x  <-  4 == 3
if (x) {
  "4 equals 3"
} else {
  "4 does not equal 3"          
}
```

As we can see, the not equal message was printed because the vector x is FALSE
```{r}
x <- 4 == 3
x
```

>__Challenge__

Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. Now do the same for 2012.

*Did anyone get a warning message like this?*
Warning in if (gapminder$year == 2012) {: the condition has length > 1 and
only the first element will be used

If your condition evaluates to a vector with more than one logical element, the function `if()` will still run, but will only evaluate the condition in the first element. Here you need to make sure your condition is of length 1.

#### any() and all()
The any() function will return TRUE if at least one TRUE value is found within a vector, otherwise it will return FALSE. This can be used in a similar way to the %in% operator. The function all(), as the name suggests, will only return TRUE if all values in the vector are TRUE.

```{r}
gapminder <-  read.csv("../data/gapminder.csv")
(any(gapminder$year==2002)) 
```
Can you solve the challenge now?

#### Key Points 
* if() { } - structure of if, then statements  
* else() { }  - structure for multiple alternate options  
* any()  - returns TRUE if any of the elements in a vector are TRUE  
* all()  - returns TRUE only if all of the elements in a vector are TRUE  

#### __Acknowledgments__
_These pages were derived in part from:Benjamin Soltoff's MACS 30500 - [Computing for the Social Sciences](https://cfss.uchicago.edu/) at University, and from [Software Carpentry](software-carpentry.org), licensed under the CC BY-NC 3.0 Creative Commons_
