---
title: "Mapping in R"
subtitle: "GEOG 693g"
author: "Amy Hessl"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Motivation
Creating a study area map is something nearly all geoscientists need to do.  There are some straightforward methods in R that make it possible to integrate with R Markdown and other programmable outlets.  Further, there is a burgeoning area of R development in spatial methods and objects that may be of interest to some.  Here we learn a few simple mapping tools.

#### Packages to Install for This Lesson:

```{r, warning=FALSE}
library("ggplot2")
theme_set(theme_bw())
library("sf")
library("rnaturalearth")
library("rnaturalearthdata")
library("rgeos")
library("ggspatial")
library("maps")
library("lwgeom")
```

### Base Map: The World
First, let us start with creating a base map of the world using ggplot2. This base map will then be extended with different map elements, as well as change the scale to focus on an area of interest. The package `rnaturalearth` provides a map of countries of the entire world. Use `ne_countries` to pull country data and choose the scale (rnaturalearthhires is necessary for scale = "large"). The function can return `sp` classes (default) or directly `sf` classes, as defined in the argument class():
```{r, warning=FALSE}
world <- ne_countries(scale = "medium", returnclass = "sf")
class(world)
```

Magically, it can be plotted with ggplot2:
```{r}
ggplot(data = world) +
    geom_sf()
```

The line ends up with a + sign, which indicates that the call is not complete yet, and each subsequent line correspond to another layer or scale. In this case, we use the geom_sf(), which  adds a geometry stored in a sf object. By default, all geometry functions use the main data defined in ggplot(), but we will see later how to provide additional data.

Note that layers are added one at a time in a ggplot call, so the *order of each layer is very important*. All data will have to be in an sf format to be used by ggplot2; data in other formats (e.g. classes from sp) will be manually converted to sf classes if necessary.

```{r}
ggplot(data = world) +
    geom_sf() +
    xlab("Longitude") + ylab("Latitude") +
    ggtitle("World map", subtitle = paste0("(", length(unique(world$name)), " countries)"))
```

#### Color and Fill
```{r}
ggplot(data = world) + 
    geom_sf(color = "black", fill = "tan")
```

We can use another colorblind friendly pallette called “viridis” for the color gradient (with option = "plasma" for the plasma variant), using the square root of the population (which is stored in the variable POP_EST of the world object):

```{r}
ggplot(data = world) +
    geom_sf(aes(fill = pop_est)) +
    scale_fill_viridis_c(option = "plasma", trans = "sqrt")
```

#### Projection and extent (coord_sf)
The function coord_sf allows to define the coordinate system, which here includes both projection and extent of the map. By default, the map will use the coordinate system of the first layer that defines one (i.e. scanned in the order provided), or if none, fall back on WGS84 (latitude/longitude, the reference system used in GPS) - which is not really a projection!

Using the argument crs, it is possible to override this setting, and project on the fly to any projection. This can be achieved using any valid PROJ4 string (here, the European-centric ETRS89 Lambert Azimuthal Equal-Area projection):

```{r}
ggplot(data = world) +
    geom_sf() +
    coord_sf(crs = "+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs ")
```
The extent of the map can also be set in coord_sf, in practice allowing to “zoom” in the area of interest, provided by limits on the x-axis (xlim), and on the y-axis (ylim). Note that the limits are automatically expanded by a fraction to ensure that data and axes don’t overlap; it can also be turned off to exactly match the limits provided with expand = FALSE:

```{r}
ggplot(data = world) +
    geom_sf() +
    coord_sf(xlim = c(-102.15, -74.12), ylim = c(7.65, 33.97), expand = FALSE)
```


#### Scale bar and North arrow: package `ggspatial`
Several packages are available to create a scale bar on a map (e.g. prettymapr, vcd, ggsn, or legendMap). We introduce here the package ggspatial, which provides easy-to-use functions.

scale_bar()  allows us to add simultaneously the north symbol and a scale bar into the ggplot map. Five arguments need to be set manually: lon, lat, distance_lon, distance_lat, and distance_legend. The location of the scale bar has to be specified in longitude/latitude in the lon and lat arguments. The shaded distance inside the scale bar is controlled by the distance_lon argument. while its width is determined by distance_lat. Additionally, it is possible to change the font size for the legend of the scale bar (argument legend_size, which defaults to 3). The North arrow behind the “N” north symbol can also be adjusted for its length (arrow_length), its distance to the scale (arrow_distance), or the size the N north symbol itself (arrow_north_size, which defaults to 6). Note that all distances (distance_lon, distance_lat, distance_legend, arrow_length, arrow_distance) are set to "km" by default in distance_unit; they can also be set to nautical miles with “nm”, or miles with “mi”.

```{r}
ggplot(data = world) +
    geom_sf() +
    annotation_scale(location = "bl", width_hint = 0.5) +
    annotation_north_arrow(location = "bl", which_north = "true", 
        pad_x = unit(0.75, "in"), pad_y = unit(0.5, "in"),
        style = north_arrow_fancy_orienteering) +
    coord_sf(xlim = c(-102.15, -74.12), ylim = c(7.65, 33.97))
```
Since that's not really a projection, the scale bar can be highly inaccurate due to changing longitudinal distances.

#### Labels...are not easy

The world data set already contains country names and the coordinates of the centroid of each country (among more information). We can use this information to plot country names, using world as a regular data.frame in ggplot2. The function geom_text can be used to add a layer of text to a map using geographic coordinates. The function requires the data needed to enter the country names, which is the same data as the world map. Again, we have a very flexible control to adjust the text at will on many aspects:

* size - the size (argument size)  
* hjust and vjust - the alignment, which is centered by default on the coordinates provided. Can be adjusted horizontally or vertically either be a number between 0 (right/bottom) and 1 (top/left) or a character (“left”, “middle”, “right”, “bottom”, “center”, “top”).  
* nudge_x and nudge_y - The text can also be offset horizontally or vertically.  
* fontface, color - The font of the text, for instance its color (argument color) or the type of font.      
* check_overlap - The overlap of labels,  which removes overlapping text.   Alternatively, when there is a lot of overlapping labels, the package `ggrepel` provides a geom_text_repel function that moves label around so that they do not overlap.  

For the text labels, we are defining the centroid of the counties with st_centroid, from the package sf. Then we combined the coordinates with the centroid, in the geometry of the spatial data frame. The package sf is necessary for the command st_centroid.  
* annotate function can be used to add a single character string at a specific location, as demonstrated here to add the Gulf of Mexico:
```{r}
library("sf")
world_points<- st_centroid(world)
world_points <- cbind(world, st_coordinates(st_centroid(world$geometry)))

ggplot(data = world) +
geom_sf() +
geom_text(data= world_points,aes(x=X, y=Y, label=name),
    color = "darkblue", fontface = "bold", check_overlap = FALSE) +
annotate(geom = "text", x = -90, y = 26, label = "Gulf of Mexico", 
    fontface = "italic", color = "grey22", size = 6) +
coord_sf(xlim = c(-102.15, -74.12), ylim = c(7.65, 33.97), expand = FALSE)
```

#### Final map
Now to make the final touches, the theme of the map can be edited to make it more appealing. We suggested the use of theme_bw for a standard theme, but there are many other themes that can be selected from (see for instance ?ggtheme in ggplot2, or the package ggthemes which provide several useful themes). Moreover, specific theme elements can be tweaked to get to the final outcome:

Position of the legend: Although not used in this example, the argument legend.position allows to automatically place the legend at a specific location (e.g. "topright", "bottomleft", etc.);
Grid lines (graticules) on the map: by using panel.grid.major and panel.grid.minor, grid lines can be adjusted. Here we set them to a gray color and dashed line type to clearly distinguish them from country borders lines;
Map background: the argument panel.background can be used to color the background, which is the ocean essentially, with a light blue;
Many more elements of a theme can be adjusted, which would be too long to cover here. We refer the reader to the documentation for the function theme.

```{r}
ggplot(data = world) + 
  geom_sf(fill= 'antiquewhite') + 
  geom_text(data= world_points,aes(x=X, y=Y, label=name), color = 'darkblue', fontface = 'bold', check_overlap = FALSE) + 
  annotate(geom = 'text', x = -90, y = 26, label = "Gulf of Mexico", fontface = 'italic', color = 'grey22', size = 6) + 
  annotation_scale(location = 'bl', width_hint = 0.5) + annotation_north_arrow(location = 'bl', which_north = 'true', pad_x = unit(0.75, 'in'), pad_y = unit(0.5, 'in'), style = north_arrow_fancy_orienteering) + 
  coord_sf(xlim = c(-102.15, -74.12), ylim = c(7.65, 33.97), expand = FALSE) +
  xlab('Longitude') + 
  ylab('Latitude') + 
  ggtitle('Map of the Gulf of Mexico and the Caribbean Sea') +
  theme(panel.grid.major = element_line(color = gray(.5), linetype = 'dashed', size = 0.5), panel.background = element_rect(fill = 'aliceblue'))
```

And if you want to save the map:  
`ggsave("map.pdf")`  
OR  
`ggsave("map_web.png", width = 6, height = 6, dpi = "screen")`  

### Study Site Map

```{r}
(sites <- data.frame(site_id = c("Coopers", "Seneca"), longitude = c(-79.7878, -79.3762), latitude = c(39.6556, 38.8348)))

```

The quickest way to add point coordinates is with the general-purpose function geom_point, which works on any X/Y coordinates, of regular data points (i.e. not geographic). As such, we can adjust all characteristics of points (e.g. color of the outline and the filling, shape, size, etc.), for all points, or using grouping from the data (i.e defining their “aesthetics”). In this example, we add the two points as diamonds (shape = 23), filled in dark red (fill = "darkred") and of bigger size (size = 4):

```{r}
ggplot(data = world) +
    geom_sf() +
    geom_point(data = sites, aes(x = longitude, y = latitude), size = 4, 
        shape = 16, color = "darkred") +
    coord_sf(xlim = c(-82, -77), ylim = c(37, 40), expand = FALSE)
```
A better, more flexible alternative is to use the power of sf: Converting the data frame to a sf object allows to rely on sf to handle on the fly the coordinate system (both projection and extent), which can be very useful if the two objects (here world map, and sites) are not in the same projection. To achieve the same result, the projection (here WGS84, which is the CRS code #4326) has to be a priori defined in the sf object:
```{r}
(sites <- st_as_sf(sites, coords = c("longitude", "latitude"), 
    crs = 4326, agr = "constant"))
```

```{r}
ggplot(data = world) +
    geom_sf() +
    geom_sf(data = sites, size = 4, shape = 16, color = "darkred") +
    coord_sf(xlim = c(-81, -73), ylim = c(36, 41), expand = FALSE) +
    geom_sf_label(data=sites, aes(label = site_id), nudge_y = 0.3) +
    coord_sf(xlim = c(-83, -77), ylim = c(37, 41), expand = FALSE)
```
Note that coord_sf has to be called after all geom_sf calls, as to supersede any former input.

#### Add State Boundaries

It would be informative to add finer administrative information on top of the previous map, starting with state borders and names. The package maps (which is automatically installed and loaded with ggplot2) provides maps of the USA, with state and county borders, that can be retrieved and converted as sf objects:
```{r}
states <- st_as_sf(map("state", plot = FALSE, fill = TRUE))
head(states)
```

State names are part of this data, as the ID variable. A simple (but not necessarily optimal) way to add state name is to compute the centroid of each state polygon as the coordinates where to draw their names. Centroids are computed with the function `st_centroid`, their coordinates extracted with `st_coordinates`, both from the package sf, and attached to the state object:

```{r}
states <- cbind(states, st_coordinates(st_centroid(states)))
head(states)
```
Note the warning, which basically says that centroid coordinates using longitude/latitude data (i.e. WGS84) are not exact, which is perfectly fine for our drawing purposes. State names, which are not capitalized in the data from maps, can be changed to title case using the function `str_to_title` from the package stringr (part of tidyverse):

```{r}
library("stringr")
states$ID <- str_to_title(states$ID)
head(states)
```

To continue adding to the map, state data is directly plotted as an additional sf layer using geom_sf. In addition, state names will be added using geom_text, declaring coordinates on the X-axis and Y-axis, as well as the label (from ID), and a relatively big font size.


```{r}
ggplot(data = world) +
    geom_sf() +
    geom_sf(data = states, fill = NA) + 
    geom_sf(data = sites, size = 4, shape = 16, color = "darkred") +
    coord_sf(xlim = c(-81, -73), ylim = c(36, 41), expand = FALSE) +
    geom_sf_label(data=sites, aes(label = site_id), nudge_x = -0.6, nudge_y = -.04) +
    coord_sf(xlim = c(-83, -77), ylim = c(37, 41), expand = FALSE)
```

#### Counties (polygon data)
County data are also available from the package maps, and can be retrieved with the same approach as for state data. This time, only counties from West Virginia are retained, and we compute their area using st_area from the package sf (depends on package lwgeom):

```{r}
counties <- st_as_sf(map("county", plot = FALSE, fill = TRUE))
counties <- subset(counties, grepl("west virginia", counties$ID))
counties$area <- as.numeric(st_area(counties))
head(counties)
```
```{r}
study_area <- ggplot(data = world) +
    geom_sf() +
    geom_sf(data = counties, fill = NA, color = gray(.7)) +
    geom_sf(data = states, fill = NA, color = gray(.5)) + 
    geom_sf(data = sites, size = 4, shape = 16, color = "darkred") +
    coord_sf(xlim = c(-81, -73), ylim = c(36, 41), expand = FALSE) +
    xlab('Longitude') + 
    ylab('Latitude') + 
    geom_sf_label(data=sites, aes(label = site_id), nudge_x = -0.6, nudge_y = -.04) +
    coord_sf(xlim = c(-82.8, -77.6), ylim = c(37, 41), expand = FALSE)
study_area
```

#### Key Points 
* There are a growing suite of packages to support GIScience in R environment.  
* Package sf is a good place to start.  
* sf elements can be plotted wiht ggplot2.  
* Elements are plotted as layers so their order is important.  
* Map geometry is an important consideration, just as it is in any GIS.  See [Overview of Coordinate Reference Systems (CRS)](https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/OverviewCoordinateReferenceSystems.pdf)

#### __Acknowledgments__
_These pages were derived in part from: [r-spatial.org](r-spatia.org)_
